---
title: "Class 12: Q13 and Q14"
author: "Renny Ng (A98061553)"
format: pdf
---
>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

```{r}
snp <- read.table("rs8067378_ENSG00000172057.6.txt")
table(snp$geno)
```
There are 108 A/AA, 233 A/G, and 121 G/G genotypes.

```{r}
AA <- snp$geno=="A/A"
AG <- snp$geno=="A/G"
GG <- snp$geno=="G/G"
```

```{r}
AAmedian <- median(snp$exp[snp$geno=="A/A"])
AAmedian
```
```{r}
AGmedian <- median(snp$exp[snp$geno=="A/G"])
AGmedian
```
```{r}
GGmedian <- median(snp$exp[snp$geno=="G/G"])
GGmedian
```

The median expression levels for each genotype are as follows:
A/A: 31.24847
A/G: 25.06486
G/G: 20.07363

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```
```{r}
ggplot(snp) + 
  aes(geno,exp, col=geno) +
  geom_boxplot() +
  xlab ("Genotype") +
  ylab ("Expression") +
  theme_bw() +
  guides(color=guide_legend("Genotypes"))
```
From this data, I infer that the relative expression of G/G is reduced compared to A/A. It appears that this SNP does indeed affect the expression of ORMDL3. 
